On the propriety of restricted Boltzmann machines

### 2016-10-19 writing/appendix-instability.Rmd
typos and refs

\begin{align*}
&\log\left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1, 1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1, 1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)} \right] = \\
& \qquad \log\left[\frac{\max\limits_{(v_1, \dots, v_V)}\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}
{\min\limits_{(v_1, \dots, v_V)}\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}\right] \\
& \qquad \ge \log\frac{\max\limits_{(v_1, \dots, v_V)}\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}
{\min\limits_{(v_1, \dots, v_V)}2^H\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)} \\
& \qquad = \max\limits_{(v_1, \dots, v_V)}\max\limits_{\boldsymbol h \in \{-1,1\}^H}\left\{\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right\} - \\
& \qquad \qquad \min\limits_{(v_1, \dots, v_V)}\max\limits_{\boldsymbol h \in \{-1,1\}^H}\left\{\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right\} -H\log2 \\
& \qquad \equiv R(\boldsymbol \theta)
\end{align*}
using above the monotonicity of $\log(\cdot)$.

>>>>

\begin{align*}
&\log\left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1, 1\}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1, 1\}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)} \right] = \\
& \qquad \log\left[\frac{\max\limits_{(v_1, \dots, v_V)}\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}
{\min\limits_{(v_1, \dots, v_V)}\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}\right] \\
& \qquad \ge \log\frac{\max\limits_{(v_1, \dots, v_V)}\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}
{\min\limits_{(v_1, \dots, v_V)}2^H\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)} \\
& \qquad = \max\limits_{(v_1, \dots, v_V)}\max\limits_{\boldsymbol h \in \{-1,1\}^H}\left\{\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right\} - \\
& \qquad \qquad \min\limits_{(v_1, \dots, v_V)}\max\limits_{\boldsymbol h \in \{-1,1\}^H}\left\{\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right\} -H\log2 \\
& \qquad \equiv R(\boldsymbol \theta_V)
\end{align*}
using above the monotonicity of $\log(\cdot)$.

### 2016-10-19 writing/draft.Rmd
typos and refs

The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs). (**TODO: deep learning, deep RBM references.**) A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information]. These models have reportedly been used with success in classification of images. (**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse (**TODO: citation**). In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical model theory and then explore the possibility of a rigorous fitting methodology. We find the RBM model class to be deficient in two fundamental ways.

@@

where $\boldsybol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathcc{R}^{V + H + V*H}$ denotes hte vector of model parameters and the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$. Let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\} = \{v_i, h_j, v_i h_j: i = 1, \dots, V, j = 1, \dots H\}$ be the vector of sufficient statistics in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$, the neg-potential function. Note that the RBM model is parameterized by $\boldsymbol \theta$containing two types of parameters, main effects and interaction effects. The main effects parameters ($\theta_{v_i}, \theta_{h_j}$) weight the values of the visible and hidden nodes and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections, or dependencies, between hidden and visible layers.

@@

**TODO: Add comparison to contrastive divergence**

>>>>

The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep, @salakhutdinov2012efficient, @srivastava2013modeling, @le2008representational for examples]. A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information]. These models have reportedly been used with success in classification of images [@larochelle2008classification, @srivastava2012multimodal]. However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast]. In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical model theory and then explore the possibility of a rigorous fitting methodology. We find the RBM model class to be deficient in two fundamental ways.

@@

where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathbb{R}^{V + H + V*H}$ denotes hte vector of model parameters and the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$. Let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\} = \{v_i, h_j, v_i h_j: i = 1, \dots, V, j = 1, \dots H\}$ be the vector of sufficient statistics in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$, the neg-potential function. Note that the RBM model is parameterized by $\boldsymbol \theta$containing two types of parameters, main effects and interaction effects. The main effects parameters ($\theta_{v_i}, \theta_{h_j}$) weight the values of the visible and hidden nodes and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections, or dependencies, between hidden and visible layers.

@@
